---
title: "forcats 0.2.0"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  fig.width = 4,
  fig.show = "hold",
  fig.asp = 0.618
)
options(tibble.print_min = 6, tibble.print_max = 6)
library(forcats)
```

forcats 0.2.0 has three new functions:

*   `as_factor()` is a generic version of `as.factor()`, which creates
    factors from character vectors ordered by appearance, rather than 
    alphabetically. This ensures means that `as_factor(x)` will always return
    the same result, regardless of the current locale.
  
*   `fct_other()` makes it easier to convert selected levels to "other":

    ```{r}
    x <- factor(rep(LETTERS[1:6], times = c(10, 5, 1, 1, 1, 1)))
    
    x %>% 
      fct_other(keep = c("A", "B")) %>% 
      fct_count()
    
    x %>% 
      fct_other(drop = c("A", "B")) %>% 
      fct_count()
    ```

*   `fct_relabel()` allows programmatic relabeling of levels:

    ```{r}
    x <- factor(letters[1:3])
    x
    
    x %>% fct_relabel(function(x) paste0("-", x, "-"))
    ```

There were a number of other minor improvements and bug fixes. You can see a full list in the [release notes](http://forcats.tidyverse.org/news/index.html#forcats-0-2-0)
